Scaffolder - Software for \change{Reproducible}{Manual} Genome Scaffolding.

### 2011-08-14 src/article.tex
Revise background section

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to
\change{not enough}{insufficient}, or multiple different overlaps between reads
and is analogous to missing pieces in the jigsaw or pieces that fit to multiple
other pieces. \pb

The advent of high\add{-}throughput sequencing methods has led to a renewed
interest in algorithms to solve the problem of genome assembly
\cite{pop2008,pop2009}. Nonetheless, due to the complexity of merging large
numbers of overlapping reads genome assembly software may be unable to produce
a complete sequence. Instead, the algorithm may generate several large
assembled regions of sequence (`contigs') composed of the many individual reads
that represent assembly into the longest possible sequences, given the data.
Nonetheless, these contigs represent a fragmented picture of the genome and
require additional work to join together into a complete sequence. \pb

The process of \change{ameliorating a draft sequence into a finished continuous
sequence}{finishing a genome sequence} can be expensive in terms of time and
laboratory effort, although the genomic data present in a set of generated
contigs may be sufficient for many research questions \cite{branscomb2002}.
Nevertheless, a continuous high-quality `finished' genome sequence does provide
a greater depth of information, such as complete resolution of repeat regions
and precise estimates of distances between genomic elements
\cite{parkhill2002,fraser2002}. The process of joining these contigs together
to form a continuous genome sequence is called the `scaffolding' or `finishing'
stage and is the focus of the software described in this article. \pb

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the known and
to-be-resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods of estimating the unknown sequence. This additional sequence is then
used to replace the gap between two contigs, joining them into a single
sequence. Once all contigs have been joined and gaps in a scaffold closed, the
genome may be considered finished.

@@

The process of resolving a genome sequence and building a scaffold uses wet
laboratory methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use the existing paired-read data from
the sequencing stage where reads were produced in pairs at a known distance
apart in the genome. The occurrence of paired reads in separate contigs can be
used to \change{probabilistically estimate}{estimate probabilistically} the
order and distance between these contigs. Alternatively, laboratory methods may
use PCR to amplify the unknown DNA in a gap region, then use traditional Sanger
sequencing to determine the sequence of this gap. Computational methods are
\remove{more} preferable as they are less costly in laboratory time and
materials compared to manual gap resolution, and use the available data that
exists following sequencing \cite{nagarajan2010}. Finally when the scaffold
cannot be completely resolved, \emph{in silico} software packages exist to
suggest the likely primers necessary for PCR amplifying the sequence in gap
regions \cite{gordon2001}. \pb

Examples of \emph{in silico} methods include using synteny to compare the
assembled contigs to a complete reference genome sequence by searching for
areas of sequence similarity between the two. Any areas of corresponding
sequence with the reference genome can be used to infer contig placement and
build the contigs into a scaffold \cite{richter2007,zhao2008,assefa2009}.
Recombination within each of the two compared genomes can, however, reduce the
effectiveness of this analysis. \pb

Repeat regions may also be responsible for multiple gaps when building
a genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms \add{which} specifically reassemble
the collapsed repeat region correctly \cite{mulyukov2002,koren2010}, while
a related approach uses unassembled sequence reads matching the regions around
a scaffold gap to construct a uniquely overlapping set of reads across it
\cite{tsai2010}. \pb

@@

These described \emph{in silico} methods provide a wide array of approaches
for merging contigs into a larger, continuous scaffold sequence. Nevertheless
the scaffolding process may still require manually inserting additional
sequences or further joining contigs using PCR-derived sequence. Moving and
editing large blocks of nucleotide text by hand however possibly introduces
human error and precludes any reproducibility of the steps in this process. \pb

The software outlined here, ``Scaffolder,'' aims to address these problems of
reproducibility by creating a file syntax and software framework for editing
a genome scaffold. Scaffolder uses a specific file format to define how
contigs are joined, additional sequences are inserted, and for the
specification of unknown regions. This syntax allows a scaffold to be updated
by simply editing the scaffold file. As such, scaffolder facilitates
a reproducible scaffolding process and provides a concise overview of how the
final genomic scaffold was constructed. \pb

>>>>

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
\remove{within the genome, }or biased and incomplete sequencing data, may
prevent the genome being assembled into a continuous sequence. This may be due
to \change{not enough}{insufficient}, or multiple different overlaps between
reads and is analogous to missing pieces in the jigsaw or pieces that fit to
multiple other pieces. \pb

The advent of high\add{-}throughput sequencing methods has led to a renewed
interest in algorithms to solve the problem of genome assembly
\cite{pop2008,pop2009}. \remove{Nonetheless, due to }The complexity of merging
large numbers of overlapping reads \add{can lead to} genome assembly software
may be unable to produce a complete sequence. Instead, the algorithm may
generate several large assembled regions of sequence (`contigs') composed of
the many individual reads\remove{ that represent assembly into the longest
possible sequences, given the data. Nonetheless, }These contigs represent
a fragmented picture of the genome and \add{therefore} require additional work
to join together into a complete sequence. \pb

The process of \change{ameliorating a draft sequence into a finished continuous
sequence}{finishing a genome sequence} can be expensive in terms of time and
laboratory effort. \change{although}{In some cases} the genomic data present in
a set of generated contigs may be sufficient for many research questions
\cite{branscomb2002}. Nevertheless, a continuous high-quality `finished' genome
sequence does provide a greater depth of information, such as complete
resolution of repeat regions and precise estimates of distances between genomic
elements \cite{parkhill2002,fraser2002}. The process of joining these contigs
together to form a continuous genome sequence is called the `scaffolding' or
`finishing' stage and is the focus of the software described in this article.
\pb

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence. \change{which}{This sequence} is useful for representing
both the known and to-be-resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods of estimating the unknown sequence. This additional sequence is
\remove{then }used to replace the gap between two contigs, joining them into
a single sequence. Once all contigs have been joined and gaps in a scaffold
closed, the genome may be considered finished.

@@

The process of \change{resolving a genome sequence and building a}{finishing
a genome} scaffold \change{may} use wet laboratory methods, \emph{in silico}
methods, or a combination of both. An example of a computational method might
use the \remove{existing} paired-read data from the sequencing stage.
\remove{where reads were produced in pairs at a known distance apart in the
genome.} The occurrence of paired reads in separate contigs can be used to
\change{probabilistically estimate}{estimate probabilistically} the order and
distance between these contigs. Alternatively, laboratory methods may use PCR
to amplify the unknown DNA in a gap region, then use \remove{traditional}
Sanger sequencing to determine the sequence of this gap. Computational methods
are \remove{more} preferable as they are less costly in laboratory time and
materials compared to manual gap resolution, and use the available data that
exists following sequencing \cite{nagarajan2010}. Finally when the scaffold
cannot be completely resolved, \emph{in silico} software packages exist to
suggest the likely primers necessary for PCR amplifying the sequence in gap
regions \cite{gordon2001}. \pb

Examples of \emph{in silico} methods include \remove{using synteny to}
comparing the assembled contigs to a complete reference genome sequence
\change{by}{to} search\remove{ing} for areas of sequence similarity between the
two. Any areas of corresponding sequence with the reference genome can be used
to infer contig placement and build the contigs into a scaffold
\cite{richter2007,zhao2008,assefa2009}. Recombination within each of the two
compared genomes can, however, reduce the efficacy of this. \pb

Repeat regions may also be responsible for multiple gaps when building
a genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms \add{which} specifically reassemble
the collapsed repeat region correctly \cite{mulyukov2002,koren2010}.
\change{while a}{A} related approach uses unassembled sequence reads matching
the regions around a scaffold gap to construct a uniquely overlapping set of
reads across \change{it}{the gap}. \cite{tsai2010}. \pb

@@

These described \emph{in silico} methods provide a wide array of approaches for
merging contigs into a larger, continuous scaffold sequence.
\remove{Nevertheless} The scaffolding process may still require manually
inserting additional sequences or further joining contigs using PCR-derived
sequence. Moving and editing large blocks of nucleotide text by hand however
\remove{possibly} introduces human error and precludes any reproducibility of
the steps in this process. \pb

The software outlined here, ``Scaffolder,'' aims to address these problems of
reproducibility by creating a file syntax and software framework for editing
a genome scaffold. Scaffolder uses a specific file format to define how
contigs are joined, additional sequences are inserted, and for the
specification of unknown regions. This syntax allows a scaffold to be updated
by simply editing the scaffold file. As such, scaffolder facilitates
a reproducible \change{scaffolding}{finishing} process and provides a concise
overview of how the final genomic scaffold was constructed. \pb
